Improving Item-Item Similarity Estimation during Collaborative Filtering

### 2014-03-24 paper/paper.tex
Add second paragraph.

...
Accurate estimations of the similarity of two users or items is valuable for a
number of data-mining applications including recommender systems, product
assortment and differentiation, as well as branding. Collaborative filtering is
often used to exploit user-item rating data to make such estimates typically by
computing the Pearson correlation or cosine similarity of vector representations
of the users or items \cite{Breese98}. Given sufficient user-item rating data,
this approach can be very effective. On the otherhand, when very little rating
data for one or both users/items is available, unmodified, these metrics can
yield dramatic discrepancies compared to values computed later once sufficient
data has been acquired. In this report, we describe two novel techniques for
estimating the true similarity of two users or items given limited common rating
data.

>>>>

...
Accurate estimations of the similarity of two users or items is valuable for a
number of data-mining applications including recommender systems, product
assortment and differentiation, as well as branding. Collaborative filtering is
often used to exploit user-item rating data to make such estimates typically by
computing the Pearson correlation or cosine similarity of vector representations
of the users or items \cite{Su2009}. Given sufficient user-item rating data,
this approach can be very effective. On the otherhand, when very little rating
data for one or both users/items is available, unmodified, these metrics can
yield dramatic discrepancies compared to values computed later once sufficient
data has been acquired. In this report, we describe two novel techniques for
estimating the true similarity of two users or items given limited common rating
data.

To keep the present work focused, we confine ourselves to memory-based
collaborative filtering (CF) in which we work with a logical user-item matrix
with elements corresponding to ratings given by a particular user to a
particular item. Further, we choose to limit our analysis to item-item
similarity but note that these techniques should be equally effective at
improving user-user similarity estimation. An item is now represented as vector
of the ratings they have received from all users. We now formalize the notion of
item-item similarity. Two alternative models for the generation of user-item
ratings are useful here in arriving at two different definitions of the {\em
true} item-item similarity.

In the first model, we imagine users are randomly presented with items to rate.
Given a sufficient period of time, all users will eventually rate all items such
that all elements of an item's vector correspond to a provided rating. The true
similarity in this model of {\em random} rating, which we denote $TrueSimR$, is
then the vector product of two item vectors. That is, 
\begin{align}
TrueSimR(A, B) = \frac{\sum\limits_{u\in U}
r_{u,A}r_{u,B}}{\sqrt{\sum\limits_{u\in U} r_{u,A}^2}
\sqrt{\sum\limits_{u\in U} r_{u,B}^2}}
\end{align}
where $U$ is the set of all users and $r_{u,I}$ is the rating given by user $u$
to item $I$.

In the second model, we imagine an unlimited supply of users each of whom
chooses to rate a subset of items. At any point in time, we may estimate the
similarity in this model of {\em preferential} rating, by computing the cosine
similarity of the two item vectors. Only in the limit of inifinite time does
this estimate converge with the true similarity, which we denote $TrueSimP$.
That is,
\begin{align}
TrueSimP(A, B) = lim_{t\to\infty}\frac{\sum\limits_{u\in U_{AB}}
r_{u,A}r_{u,B}}{\sqrt{\sum\limits_{u\in U_A} r_{u,A}^2}
\sqrt{\sum\limits_{u\in U_B} r_{u,B}^2}}
\end{align}
where $U_A$ is the set of users who have rated $A$, $U_B$ is the set of users
who have rated $B$, and $U_{AB}$ is the set of users who have rated both $A$ and
$B$.
